Quantifying the manageability of pollination networks in an invasion context

### 2017-04-26 paper/response_to_reviewers.Rmd

---
title: "Response to comments ELE-00205-2017"
output: pdf_document
csl: ecology-letters.csl
classoption: a4paper
bibliography: references.bib
header-includes:
- \usepackage{setspace}
---

@@

\onehalfspacing

Dear Dr Jonathan Chase,

We are thankful for the opportunity to resubmit the manuscript for consideration. 
In the response to the handling editor section below, we have endeavoured to explain how our approach does still provide ecological insight despite the assumption of linearity. 
To aid in the review process we have included in this document as well the response to the reviewers' and editor's comments of the previous submission (ELE-00738-2016).

Before we dive into the responses, we would like to remark that the major concern identified by the new handling editor—the assumption of linearity—was not overlooked in the previous round of review. 
In the initial version of the manuscript, we mainly focus on details of the method and the study system and only mention in passing that our approach is based on a linearization of the network of interactions. 
Reviewer 3 and 1 in particular, did not find this to be a fundamental problem, although they encouraged us to be explicit about this assumption and thoroughly discuss the implications of this simplification. 
In hindsight, perhaps we focused executively on the potential limitations of our approach and did a poor job justifying its advantages.

We hope this response and the modifications made to the main text are convincing enough for the manuscript to be reconsidered for publication at Ecology Letters.

****

# Response to comments ELE-00738-2016

## To the handling editor

### 2017-05-01 paper/response_to_reviewers.Rmd

---
title: "Response to comments ELE-00205-2017"
output: pdf_document
csl: ecology-letters.csl
classoption: a4paper
bibliography: references.bib
header-includes:
- \usepackage{setspace}
---

\doublespacing

Dear Dr Jonathan Chase,

We are thankful for the opportunity to resubmit the manuscript for consideration. 
In the response to the handling editor section below, we have endeavoured to explain how our approach does still provide ecological insight despite the assumption of linearity. 
To aid in the review process we have included in this document as well the response to the reviewers' and editor's comments of the previous submission (ELE-00738-2016) and have updated the references to line numbers when appropriate.

@@

Before we dive into the responses, we would like to remark that the major concern identified by the new handling editor—the assumption of linearity—was not overlooked in the previous round of review. 
In the initial version of the manuscript, we mainly focus on details of the method and the study system and only mention in passing that our approach is based on a linearization of the network of interactions. 
Reviewer 3 and 1 in particular, did not find this to be a fundamental problem, although they encouraged us to be explicit about this assumption and thoroughly discuss the implications of this simplification. 
In hindsight, perhaps we focused executively on the potential limitations of our approach and did a poor job justifying its advantages.

We hope this response and the modifications made to the main text are convincing enough for the manuscript to be reconsidered for publication at Ecology Letters.

## Response to the handling editor

We agree with the handling editor that the most notable limitation of our approach is that it is based on a linear approximation of the dynamics of the system. 
Naturally, it is important to take this limitation into account when applying the framework. 
We disagree, however, with the conclusion that this limitation renders the approach ecologically irrelevant.

We approach the editor's comments in two stages. 
First, we argue that linear models can be useful to describe the dynamics of ecological communities, and pollination communities in particular. 
Additionally, we believe that the editor's observations potentially arises from a misunderstanding of the scope and goals of our study.
Therefore, we then provide some further clarification in this regard and to structural controllability. 
Finally, we provide a detailed account of the editions we have done to the manuscript in order to prevent this kind of misunderstandings.

While ecological systems are inherently non-linear, variations of Lokta-Volterra and other linear response models have been widely and recently used in ecology to describe both multitrophic ecological communities [@Heckmann2012; @Heath2014] and mutualistic pollination networks [@Lever2014; @Rohr2014; @Valdovinos2016]. 
Importantly several studies have found that, despite the lack of realism, parametrizations with linear responses can provide a good starting point to investigate community stability (ref), population size [@Mackinson2003], and relative species importance [@Zhao2016]. 
Similarly, we view the structural controllability as a *starting point* to investigate the way ecological communities could be steered from an equilibrium point to another in an informed fashion.

Furthermore, the editor points out that a structural controllability approach would not be useful in ecological systems because it will be "only valid in an infinitesimal border" around an initial equilibrium point. 
This comment would be completely appropriate if the local linear approximation at the equilibrium point were used to *design* the—inherently non-local—control trajectory that drives the system to an alternative state. 
Our manuscript, however, has a different aim. 
We use the structure of interaction network and determine the set of possible "control interventions" to gain network and species-level ecological that could be used in management settings.

In our manuscript, a control intervention is a set of species with which full control of the community could be gained. 
The control interventions we study are optimal from a structural point of view.
However, in practice, some of these interventions would be more appropriate than others depending on several factors like the starting and end state of the community, the feasibility of modifying the abundance of some species, the available resources, and naturally the particular dynamics of the system. 
As the behaviour of the community is governed by both structure and dynamics, by extension, our approach is ill suited to design a specific control trajectory. 
Again, we emphasise that this was not our objective and that the development of numerical approaches like the one proposed by @Cornelius2013 might be more promising in this regard.

Nevertheless, we see the fact that we focus on structure as and advantage rather than a disadvantage. 
As one of the previous reviewers highlighted, our manuscript "...presents to ecologists a novel way of studying the relationship between topology and dynamics that is less dependent (when compared to other methods) on particular choices about how dynamics are modelled". 
It can, therefore, be used to gain insight from different ecological systems and is readily applicable to currently available data. 
Furthermore, by focusing on the network structure, the conclusions obtained will be locally valid, not only close to equilibrium points but anywhere as long as the species interactions remain.

Specifically, we made the following modifications to the manuscript. 
In the intro we ...
As before, in the methods section, we start by stating the assumption of linearity (Lines XXX). 
However, we now also explicitly explain how this assumption does not invalidate the gained insight (Lines XXX). 
Finally, in the discussion we ...

### 2017-05-01 paper/response_to_reviewers.Rmd

---
title: "Response to comments ELE-00205-2017"
output: pdf_document
csl: ecology-letters.csl
classoption: a4paper
bibliography: references.bib
header-includes:
- \usepackage{setspace}
---

\doublespacing

Dear Dr Jonathan Chase,

We are thankful for the opportunity to resubmit our manuscript for consideration. 
In our response to the handling editor section (below), we have endeavoured to explain how our approach still provides ecological insight despite the assumption of linearity. 
To aid in the review process, we have included the response to the reviewers' and editor's comments of the previous submission (ELE-00738-2016) in this document as well and have updated the references to line numbers when appropriate.

Before we dive into the detailed responses, we would like to draw your attention to the fact that the major concern identified by the new handling editor—the assumption of linearity—was not actually overlooked in the previous round of review. 
In the initial version of the manuscript, we mainly focused on details of the method and the study system and only mentioned in passing that our approach is based on a linearization of the network of interactions. 
Reviewers 3 and 1 in particular did not find this to be a fundamental problem; nevertheless they encouraged us to be explicit about this assumption and thoroughly discuss the implications of this simplification. 
In hindsight, perhaps these changes focused too intensely on the potential limitations of our approach and did a poorer job of justifying its advantages.

### 2017-05-02 paper/response_to_reviewers.Rmd

---
title: "Response to comments ELE-00205-2017"
output: pdf_document
csl: ecology-letters.csl
classoption: a4paper
bibliography: references.bib
header-includes:
- \usepackage{setspace}
---

\doublespacing

Dear Dr Jonathan Chase,

We are thankful for the opportunity to resubmit our manuscript for consideration. 
In our response to the handling editor section (below), we have endeavoured to explain how our approach still provides ecological insight despite the assumption of linearity. 
To aid in the review process, we have included the response to the reviewers' and editor's comments of the previous submission (ELE-00738-2016) in this document as well and have updated the references to line numbers when appropriate.

Before we dive into the detailed responses, we would like to draw your attention to the fact that the major concern identified by the new handling editor—the assumption of linearity—was not actually overlooked in the previous round of review. 
In the initial version of the manuscript, we mainly focused on details of the method and the study system and only mentioned in passing that our approach is based on a linearization of the network of interactions. 
Reviewers 3 and 1 in particular did not find this to be a fundamental problem; nevertheless, they encouraged us to be explicit about this assumption and thoroughly discuss the implications of this simplification. 
In hindsight, perhaps these changes focused too intensely on the potential limitations of our approach and did a poorer job of justifying its advantages.

@@

We hope this response and the modifications made to the main text are convincing enough for the manuscript to be reconsidered for publication at Ecology Letters.

## Response to the handling editor

We agree with the handling editor that the most notable assumption of our approach is that it is based on a linear approximation of the dynamics of the system. 
Naturally, it is important to take this into account when applying the framework. 
We disagree, however, with the conclusion that this detail renders the approach ecologically irrelevant.

We will approach the editor's comments in two stages below. 
First, we will argue that linear models can be useful to describe the dynamics of ecological communities and pollination communities in particular. 
Second, we believe that the editor's observations potentially arises from our failure to clearly delineate the scope and goals of our study.
Therefore, we then provide some further clarification in this regard and to the goals of structural controllability in general. 
Finally, we provide a detailed account of the edits we have made in the manuscript in order to prevent this kind of misunderstandings going forward.

While ecological systems are inherently non-linear, variations of Lokta-Volterra and other models with linear functional response have been widely used in ecology to describe both multitrophic ecological communities [@Heckmann2012; @Heath2014] and mutualistic pollination networks [@Lever2014; @Rohr2014; @Valdovinos2016]. 
Importantly several studies have found that despite the lack of "realism", parametrizations with linear responses still provide a valuable starting point to investigate community stability [@Allesina2012], population size [@Mackinson2003], and relative species importance [@Zhao2016]. 
Similarly, we view the structural controllability as a *starting point* to investigate the way ecological communities could be steered from an equilibrium point to another in an informed fashion.

An ecological system near equilibrium can be linearised by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$. 
Here, $\mathbf{B}$ is the "input matrix" and $u(t)$ is the input or control vector. 
The matrix $\mathbf{B}$ has size $S \times D$, where $S$ is the number of species in the community and $D$ is the number of species whose abundances are modified by the input vector. 
The editor points out that a structural controllability approach would not be useful in ecological systems because it will be "only valid in an infinitesimal border" around an initial equilibrium point. 
This comment would be completely appropriate if our aim were to *design the control trajectory* that drives the system from an equilibrium point to an alternate stable state. 
In other words, it would be inappropriate to design $u(t)$ based exclusively a local linear approximation at the initial equilibrium point. 
Our manuscript, however, has a completely different aim.

We use the structure of interaction network to determine the set of nodes that could be used to control the system. 
This is, we use information contained in $\mathbf{A}$ to determine $\mathbf{B}$ (and by extension $D$). 
Our most novel contributions are to use $D$ and knowledge of all the possible ways in which $\mathbf{B}$ can be parameterised in order to gain insight of the controllability of the network and role of the species that compose it.
Understanding $\mathbf{B}$ is a prerequisite to designing $u(t)$, regardless of whether we are interested on controllability near an equilibrium point or along a linearisation of the system's nominal trajectory outside of equilibrium.
Even existing numerical approaches that operate in a fully nonlinear setting like those proposed by @Cornelius2013 require previous knowledge of the set of nodes (species) to which to apply the control signal.

We see the fact that we focus on $\mathbf{B}$ and the network structure as an advantage rather than a disadvantage. 
As one of the previous reviewers highlighted, our manuscript "...presents to ecologists a novel way of studying the relationship between topology and dynamics that is less dependent (when compared to other methods) on particular choices about how dynamics are modelled". 
It can, therefore, be used to gain insight from different ecological systems and is readily applicable to currently available data.

As it has been acknowledged in the control theory literature, "while many complex systems are characterised by nonlinear interactions between the components, the first step in any control challenge is to establish the controllability of the locally linearized system... Before we can explore the fully nonlinear dynamical setting, which is mathematically much harder, we must understand the impact of the topological characteristics on linear controllability, serving as a prerequisite of nonlinear controllability" [@Liu2016].

Along similar lines to the above, we made the following modifications to the manuscript. 
In the intro we ...
As before, in the methods section, we start by stating the assumption of linearity (Lines XXX). 
However, we now also explicitly explain how this assumption does not invalidate the gained insight (Lines XXX). 
Finally, in the discussion we ...

### 2017-05-02 paper/response_to_reviewers.Rmd

---
title: "Response to comments ELE-00205-2017"
output: pdf_document
csl: ecology-letters.csl
classoption: a4paper
bibliography: references.bib
header-includes:
- \usepackage{setspace}
---

\doublespacing

Dear Dr Jonathan Chase,

We are thankful for the opportunity to resubmit our manuscript for consideration. 
In our response to the handling editor (below), we have endeavoured to explain how our approach still provides ecological insight despite the assumption of linearity. 
To aid in the review process, we have included the response to the reviewers' and editor's comments of the previous submission (ELE-00738-2016) in this document as well and have updated the references to line numbers when appropriate.

Before we dive into the detailed responses, we would like to draw your attention to the fact that the major concern identified by the new handling editor—the assumption of linearity—was not actually overlooked in the previous round of review. 
In the initial version of the manuscript, we mainly focused on details of the method and the study system and only mentioned in passing that our approach is based on a linearization of the network of interactions. 
Reviewers 3 and 1 in particular did not find this to be a fundamental problem; nevertheless, they encouraged us to be explicit about this assumption and thoroughly discuss the implications of this simplification. 
In hindsight, perhaps these changes focused too intensely on the potential limitations of our approach and did a poorer job of justifying its sustained advantages.

@@

We hope this response and the modifications made to the main text are convincing enough for the manuscript to be reconsidered for publication at Ecology Letters.

## Response to the handling editor

We agree with the handling editor that the most notable assumption of our approach is that it is based on a linear approximation of the dynamics of the system. 
Naturally, it is important to take this into account when applying the framework. 
We disagree, however, with the conclusion that this detail renders the approach ecologically irrelevant.

We will approach the editor's comments in two stages below. 
First, we will argue that linear models can be useful to describe the dynamics of ecological communities and pollination communities in particular. 
Second, we believe that the editor's observations potentially arises from our failure to clearly delineate the scope and goals of our study.
Therefore, we then provide some further clarification in this regard and to the aims of structural controllability in general. 
Finally, we provide a detailed account of the edits we have made in the manuscript in order to prevent this kind of misunderstanding going forward.

While ecological systems are inherently non-linear, variations of Lokta-Volterra and other models with linear functional responses have been widely used in ecology to describe both multitrophic ecological communities [@Heckmann2012; @Heath2014] and mutualistic pollination networks [@Lever2014; @Rohr2014; @Valdovinos2016]. 
Importantly, several studies have found that, despite the lack of "realism", parametrizations with linear responses still provide a valuable starting point to investigate community stability [@Tang2014], population size [@Mackinson2003], and relative species importance [@Zhao2016]. 
Similarly, we view the structural controllability as a *starting point* to investigate the way ecological communities could be steered from an equilibrium point to another in an informed fashion.

As noted in our manuscript, the dynamics of an ecological system near equilibrium can be represented by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$. 
Here, $\mathbf{B}$ is the "input matrix" and $u(t)$ is the control vector. 
The matrix $\mathbf{B}$ has size $S \times D$, where $S$ is the number of species in the community and $D$ is the number of species whose abundances are modified by the control vector. 
The editor points out that a structural-controllability approach would not be useful in linear ecological systems because it will be "only valid in an infinitesimal border" around an initial equilibrium point. 
This comment would be completely appropriate if our aim were to design the time-varying control trajectory that drives the system from one equilibrium point another. 
In other words, it would be inappropriate to design $u(t)$ based exclusively on a local linear approximation at the initial equilibrium.  
Our manuscript, however, has a completely different aim.

We instead use the structure of interaction network to determine the set of nodes that could be used to control the system. 
That is, we use the information contained in $\mathbf{A}$ to generate a supportable estimate of $\mathbf{B}$ (and by extension $D$). 
Our most novel contributions are to use $D$ and knowledge of all the possible ways in which $\mathbf{B}$ can be structured in order to gain insight of the controllability of the network and the role of the species that compose it.
Understanding $\mathbf{B}$ is a prerequisite to designing $u(t)$, regardless of whether we are interested in controllability near an equilibrium point or along a linearisation of the system's nominal trajectory outside of equilibrium.
Indeed, even existing numerical approaches that operate in a fully nonlinear setting like those proposed by @Cornelius2013 require previous knowledge of the set of nodes (species) to which to apply the control signal.

Furthermore, we see the fact that we focus on $\mathbf{B}$ and the network structure as an advantage rather than a disadvantage. 
As one of the previous reviewers highlighted, our manuscript "...presents to ecologists a novel way of studying the relationship between topology and dynamics that is less dependent (when compared to other methods) on particular choices about how dynamics are modelled". 
It can, therefore, be used to gain insight from different ecological systems and is readily applicable to currently available data.

### 2017-05-03 paper/response_to_reviewers.Rmd

We agree with the handling editor that the most notable assumption of our approach is that it is based on a linear approximation of the dynamics of the system. 
Naturally, it is important to take this into account when applying the framework. 
We disagree, however, with the conclusion that this detail renders the approach ecologically irrelevant.

We will approach the editor's comments in two stages below. 
First, we will argue that linear models can be useful to describe the dynamics of ecological communities and pollination communities in particular. 
Second, we believe that the editor's observations potentially arises from our failure to clearly delineate the scope and goals of our study.
Therefore, we then provide some further clarification in this regard and to the aims of structural controllability in general. 
Finally, we provide a detailed account of the edits we have made in the manuscript in order to prevent this kind of misunderstanding going forward.

While ecological systems are inherently non-linear, variations of Lokta-Volterra and other models with linear functional responses have been widely used in ecology to describe both multitrophic ecological communities [@Heckmann2012; @Heath2014] and mutualistic pollination networks [@Lever2014; @Rohr2014; @Valdovinos2016]. 
Importantly, several studies have found that, despite the lack of "realism", parametrizations with linear responses still provide a valuable starting point to investigate community stability [@Tang2014], population size [@Mackinson2003], and relative species importance [@Zhao2016]. 
Similarly, we view the structural controllability as a *starting point* to investigate the way ecological communities could be steered from an equilibrium point to another in an informed fashion.

As noted in our manuscript, the dynamics of an ecological system near equilibrium can be represented by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$. 
Here, $\mathbf{B}$ is the "input matrix" and $u(t)$ is the control vector. 
The matrix $\mathbf{B}$ has size $S \times D$, where $S$ is the number of species in the community and $D$ is the number of species whose abundances are modified by the control vector. 
The editor points out that a structural-controllability approach would not be useful in linear ecological systems because it will be "only valid in an infinitesimal border" around an initial equilibrium point. 
This comment would be completely appropriate if our aim were to design the time-varying control trajectory that drives the system from one equilibrium point another. 
In other words, it would be wrong to design $u(t)$ based exclusively on a local linear approximation at the initial equilibrium. 
Our manuscript, however, has a completely different aim.

We instead use the structure of interaction network to determine the set of nodes that could be used to control the system. 
That is, we use the information contained in $\mathbf{A}$ to generate a supportable estimate of $\mathbf{B}$ (and by extension $D$). 
Our most novel contributions are to use $D$ and knowledge of all the possible ways in which $\mathbf{B}$ can be str uctured in order to gain insight of the controllability of the network and the role of the species that compose it.
Understanding $\mathbf{B}$ is a prerequisite to designing $u(t)$, regardless of whether we are interested in controllability near an equilibrium point or along a linearisation of the system's nominal trajectory outside of equilibrium.
Indeed, even existing numerical approaches that operate in a fully nonlinear setting like those proposed by @Cornelius2013 require previous knowledge of the set of nodes (species) to which to apply the control signal.

### 2017-05-03 paper/manuscript.Rmd
update manuscript to address handling editor concerns

Conveniently the information necessary to determine the minimum number of driver nodes *D* in a network with linear dynamics is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
We therefore start here from the assumption that an ecological network near equilibrium points can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example the species' abundances), an external time-varying input *u(t)*, and two matrices *A* and *B*, which encapsulate information about the network structure and how the species respond to the external input, respectively. 
Even though linearity might be a strong assumption, it has been shown that fundamental insight about the control of complex systems can be gained without a detailed knowledge of the nonlinear dynamics and the system parameters [@Liu2016], an important step towards ultimately understanding the controllability of systems with nonlinear dynamics [@Liu2011]. 
This means that structural controllability can be applied to a wide range of network representations of ecological communities to provide an indication of our ability to control them. 
Moreover, we show here how quantitative data about species interactions can help us move past some of the limitations of structural controllability by better approximating the role of interspecific dynamic relationships without being overly dependent on the choices of how these dynamics are modelled or characterised.

@@

In this study, we illustrate how a control-theoretic approach can be better adopted in network ecology to evaluate the effect of invasions and other kinds of perturbations. 
Although our pollination specific results might not be directly translatable to other ecological systems because of the degree constraints imposed by bipartite networks, the approach we propose is applicable wherever species abundances are influenced by their interactions, and exciting open questions lie ahead. How to design the precise "control signals" to reach a desired ecosystem state or conservation outcome? 
What are the implications of assuming linear species dynamics?
How important it is to include several interaction types for our understanding of manageability and species importance?
What are the implications for species coexistence? 
Which are the trade-offs between persistence at the species and the community level? 
Answering each of these questions might require its evaluation in different ecological systems, an explicit integration of control theory with numerical models of species densities [@Cornelius2013; @Gibson2016], and experimental tests on simple communities. 
Nevertheless, the potential rewards are encouraging from both an ecological and conservation perspective, where an integrated approach can shift our focus beyond the identification of ideal targets for intervention to the design of informed interventions that legitimately achieve restoration goals.

>>>>

Conveniently the information necessary to determine the minimum number of driver nodes *D* in a system with linear dynamics is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example the species' abundances), an external time-varying input *u(t)* (the control vector), and two matrices *A* and *B*, which encapsulate information about the network structure and how the species respond to the external input, respectively. 
The matrix *B* has size $S \times D$, where *S* is the number of species in the community. 
Briefly, the focus of structural controllability, which we employ here, is to use the information contained in $\mathbf{A}$ to generate a supportable estimate of $\mathbf{B}$ (and by extension $D$).

This focus allows us to gain insight of the controllability of the network and the role of the species that compose it without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off is that, because of the assumption that ecological communities can be modeled with linear functional responses, at least near equilibrium points, structural controllability is not adequate to design a the time varying control signal *u(t)* that can drive the system from one particular equilibrium to another.	
Nevertheless, the lessons gained when assuming linearity—both at the network and the species level—are a prerequisite for understanding nonlinear control [@Liu2011; @Liu2016].

@@

In this study, we illustrate how a control-theoretic approach can be better adopted in network ecology to evaluate the effect of invasions and other kinds of perturbations. 
We study the different ways a management intervention can be structured and provide a starting point for the study controllability of ecological systems.
Although our pollination specific results might not be directly translatable to other ecological systems because of the degree constraints imposed by bipartite networks, the approach we propose is applicable wherever species abundances are influenced by their interactions, and exciting open questions lie ahead. 
How to design the precise "control signals" to reach a desired ecosystem state or conservation outcome? 
What are the implications of assuming fully nonlinear dynamics?
How important it is to include several interaction types for our understanding of manageability and species importance?
What are the implications for species coexistence? 
Which are the trade-offs between persistence at the species and the community level? 
Answering each of these questions might require its evaluation in different ecological systems, an explicit integration of control theory with numerical models of species densities [@Cornelius2013; @Gibson2016], and experimental tests on simple communities. 
Nevertheless, the potential rewards are encouraging from both an ecological and conservation perspective, where an integrated approach can shift the focus beyond the identification of ideal targets for intervention to the design of informed interventions that legitimately achieve restoration goals.
